### 2017-07-07 motivation.tex
edits

We present our ideas in the context of our DSL embedded in OCaml,
which offers a \C{DB} monad to define and compose database
computations given in terms of SQL operations.  Besides the usual
\C{bind} and \C{return} operators, the monad offers an \C{atomically}
combinator that executes a database computation as an atomic
transaction and returns the result.

@@

Fig.~\ref{fig:new_order_code} shows a simplified version of the TPC-C
\C{new\_order} transaction written in this language. TPC-C is a
widely-used Online Transaction Processing (OLTP) benchmark that models
an order-processing system for a wholesale parts supply business. The
business logic is captured in 5 database transactions that operate on
9 tables; \C{new\_order} is one such transaction that uses
\C{District}, \C{Order}, \C{New\_order}, \C{Stock}, and
\C{Order\_line} tables. The transaction acts on the behalf of a
customer, whose id is \C{c\_id}, to place a new order for a given
set of items (\C{item\_reqs}), to be served by a warehouse under the
district identified by \C{d\_id}.  Fig.~\ref{fig:schema} illustrates
the relationship among these different tables.

The transaction manages order placement by invoking appropriate SQL
functionality, captured by various calls to functions defined by the
\C{SQL} module. All \C{SQL} functions take the table name (a nullary
constructor) as their first argument. The higher-order \C{SQL.select1}
function accepts a boolean function that describes the selection
criteria, and returns any record that meets the criteria (it models
the SQL query \C{SELECT \ldots\xspace LIMIT 1}). \C{SQL.update} also
accepts a Boolean function (its 3$^{rd}$ argument) to select the records to be
updated. Its 2$^{nd}$ argument is a function that maps each selected
record to a new (updated) record. \C{SQL.insert} inserts a given
record into the specified table in the database.

@@

\begin{figure}[t]
  \centering
	\begin{subfigure}{0.48\textwidth}
		\includegraphics[width=0.9\textwidth]{Figures/schema1}
    \caption{A valid TPC-C database. The only existing order belongs
      to the district with \C{d\_id}=11. It's id (\C{o\_id}) is one
      less than the district's \C{d\_next\_o\_id}, and it's order
      count (\C{o\_ol\_cnt}) is equal to the number of order line
      records whose \C{ol\_o\_id} is equal to the order's id.  }
		\label{fig:tpcc_db1}
	\end{subfigure}
	\begin{subfigure}{0.48\textwidth}
		\includegraphics[width=0.9\textwidth]{Figures/schema2}
    \caption{The database in Fig.~\ref{fig:tpcc_db1} after correctly
      executing a \C{new\_order} transaction. A new order record is
      added whose \C{o\_id} is equal to the \C{d\_next\_o\_id} from
      Fig.~\ref{fig:tpcc_db1}. The district's \C{d\_next\_o\_id} is
      incremented. The order's \C{o\_ol\_cnt} is 2, reflecting the
      actual number of order line records whose \C{ol\_o\_id} is equal
      to the order's id (2).}
		\label{fig:tpcc_db2}
	\end{subfigure}

@@

Because the aforementioned reasoning is reasonably simple to perform
manually, verifying the soundess of TPC-C's consistency conditions
would appear to be feasible.  Serializability aids the tractability of
verification by preventing any interference among concurrently
executing transactions while the \C{new\_order} transaction executes.
Under weak isolation\footnote{Weak isolation does not violate
  atomicity as long as the witnessed effects are those of committed
  transactions}, however, interferences of various kinds are
permitted.  Although the verification problem for weakly isolated
transactions would appear to be superficially similar to the
verification of (racy) concurrent programs (e.g., garbage
collectors~\cite{JLP+14,GHE15,HPQ+15}), weak isolation introduces new
challenges arising from the use of transactions, and new opportunities
arising from the fact that the store abstraction used by transactions
is a relational database, not low-level memory.

To illustrate some of these challenges, consider the behavior of the
\C{new\_order} transaction when executed with a \emph{Read Committed}
(RC) isolation level, the default isolation level in 8 of the 18
databases studied in~\ref{bailishotos}.  An RC transaction is
isolated from \emph{dirty writes}, i.e., writes of uncommitted
transactions, but is allowed to witness the writes of concurrent
transactions as soon as they are committed. Thus, with two concurrent
instances of the \C{new\_order} transaction (call them $T_1$ and
$T_2$), both concurrently placing new orders for different customers
under the same district (\C{d\_id}), RC isolation allows the
execution shown in Fig.~\ref{fig:new_order_execs}.

\begin{figure}[!t]
\includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
\caption{\small An RC execution involving two instances ($T_1$ and
  $T_2$) of the \C{new\_order} transaction depicted in
  Fig.~\ref{fig:new_order_code}. 
  Both instances read the \C{d\_id} \C{District} record concurrently,
  because neither transaction is committed when the reads are
  executed.  The subsequent operations are effectively sequentialized,
  since $T_2$ commits. Nonetheless, reading the same value for
  \C{d\_next\_o\_id} prompts both instances to add \C{Order} records
  with same ids, which inturn triggers the the violation of TPC-C's
  consistency condition.}
\label{fig:new_order_exec}
\end{figure}

@@

Notably, this example does not exhibit any of the anomalies that
characterize RC isolation~\cite{berenson}. For instance, there are no
\emph{lost writes}; both concurrent transactions' writes are present
in the final state of the database. Thus, program analyses that aim to
determine appropriate isolation by checking for possible
manifestations of these anomalies would fail to identify grounds for
promoting the isolation level of \C{new\_order} to something stronger.
Yet, if we take the semantics of the application into account, it is
quite clear that RC is not an appropriate isolation level for
\C{new\_order}.

While reasoning in terms of anomalies is cumbersome and inadequate,
reasoning about weak isolation in terms of low-level
traces~\cite{adyaphd,gotsmanconcur15} on memory read and write actions
complicates high-level reasoning.  A possible alternative would
interleave weak isolation implementation details within the program,
yielding a (more-or-less) conventional concurrent program that can be
then subject to classical concurrent verification methods.
Considering the size and complexity of real-world transaction systems,
this strategy is unlikely to scale.

In this paper, we adopt a different approach that \emph{lifts}
isolation semantics (\emph{not} their implementations) to the
application layer, providing a principled framework to simultaneously
reason about application invariants and isolation properties.  To
illustrate this idea informally, consider how we might verify that
\C{new\_order} is sound when executed under \emph{Snapshot Isolation}
(SI), and isolation level stronger than RC. Snapshot isolation
allows transactions to be executed against their respective private
snapshots of the database, thus admitting concurrency, but it also
requires that there not be any write-writeconflicts among
concurrent transactions. Write-write conflicts can be eliminated in various
ways; either through detection followed by a rollback, or through
exclusive locks, or a combination of both, leading to different ways
in which SI can be realized. For instance, one possible implementation
of SI, close to the one use by PostgreSQL~\cite{postgres-ssi},
executes a transaction against its private snapshot of the database,
but obtains exclusive locks on the actual records in the database
before performing writes. A write is performed only of the record
hasn't already been updated by a concurrent transaction, i.e,  when it
doesn't result in a write-write conflict.  Otherwise, the transaction is
rolledback. This implementation is demonstrated for a simple database
with three records, $a$, $b$, and $c$ in Fig.~\ref{fig:RR-postgres}.

\begin{figure}[t]
\includegraphics[scale=0.5]{Figures/RR-postgres}
\caption{Database state transitions corresponding to an execution of
  an SI transaction $T_1$ on a PostgreSQL-like store. $T_2$ and $T_3$
  are concurrent transactions. }
\label{fig:rr-postgres}
\end{figure}
\begin{figure}[t]
\includegraphics[scale=0.5]{Figures/RR-abstract}
  \caption{An abstract execution that includes the concrete execution
  shown in Fig.~\ref{fig:rr-postgres}. It has no locks or snapshots,
  admits fewer interleavings, yet results in the same post-state.  }
\label{fig:rr-abstract}
\end{figure}

Implementations of SI on real databases (e.g., PostgreSQL) are
complicated, often running into thousands of lines of code.
Nonetheless, the semantics of Snapshot Isolation, in terms of how it
effects the transitions on the database state, can be captured in a
fairly simple model. Firstly, although SI admits concurrent
transactions, their effects are  not witnessed in the current
transaction (call it $T$), during $T$'s execution. Thus, insofar as
$T$ is concerned, as long as it executes, the global state doesn't
change. More formally, for every consecutive pair of global states
$(\stg,\stg')$ witnessed by $T$ during its execution, $\stg'=\stg$.
After $T$ finishes execution, it commits to the actual database, which
incorporates the effects of concurrent transactions. In executions
where $T$ successfully commits, concurrent transactions are guaranteed
to not be in write-write conflict with $T$. Thus, if $\stg$ is the
global state that $T$ witnessed when it finished execution (which
would be same as the snapshot state), and $\stg'$ is the state to
which $T$ commits, then the diff between $\stg$ and $\stg'$ shouldn't
be in a write-write conflict with $T$. To concretize this notion, let
the database state be a map from transaction variables to values, and
let $\stl$ denote a transaction-local log that maps the variables
being written to their updated values. Then the absence of write-write
conflicts between $T$ and the diff between $\stg$ and $\stg'$ can be
defined as: $\forall x\in\mathit{dom}(\delta)$, $\stg'(x) = \stg(x)$.
To summarize, the semantics of SI can be captured as an axiomatization
over transitions of the database state ($\Delta \longrightarrow
\Delta'$) during the lifetime an SI transaction ($T$):
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.
\end{itemize}

@@

This simple characterization of SI isolation allows us to verify the
consistency conditions associated with the \C{new\_order} transaction.
First, since the database doesn't change ($\Delta' = \Delta$) during
the execution of the transaction's body, we can reason about
\C{new\_order} as though it executed in complete isolation until its
commit point, leading to a verification process similar to what would
been applied when reasoning about serializability.  When
\C{new\_order} finishes execution, but before it commits, the SI
axiomatization shown above requires us to consider global state
transitions $\stg \longrightarrow \stg'$ that do not include changes
to the records ($\stl$) written by \C{new\_order}, i.e.,
$\forall(x\in\delta).~\Delta'(x) = \Delta(x)$. Since concurrent
\C{new\_order} transactions that write to the same \C{District} record
modify the record by incrementing its \C{d\_next\_o\_id} field, their
transitions do not satisfy the necessary condition imposed by the SI
axiomatization, allowing us to rightfully ignore their interference.
This leaves the interference due to concurrent \C{new\_order}
transactions that write to a different district record.  Fortunately,
such interference does not prompt \C{new\_order} to violate the
invariant, which can be established by adopting a reasoning similar to
the one used on concurrent program logics, such as the Rely-Guarantee
logic, customized for the database programs.

@@

The second observation that informs our approach is one that pertains
to automation. Program verification, even when machine-aided, often
entails significant annotation burden in the form of intermediary
assertions and loop invariants required to prove a program correct.
This is certainly true for concurrent program logics, such as
Rely-Gurantee, which extend Hoare logic with additional artifacts and
where (stable) intermediary assertions and loop invariants remain a
major source of annotation burden.  However, a relational database is
a significantly simpler abstraction than shared memory. There are no
pointers, linked data structures, or aliasing.  Although a database
essentially abstracts a mutable state, the state is mutated through a
well-defined fixed number of interfaces (SQL statements), each tagged
with a logical formula describing what records are accessed and
updated.

This observation leads us away from thinking of database transactions
as concurrent imperative programs.  Instead, we see value in viewing
them as essentially functional computations that manage database state
monadically. We find it useful to reason about
statements that mutate the database state, not in terms of a pre- and
post-condition pair, but in terms of a state transformer that relates
the pre- and post-states of a statement. This state transformer
semantics can be defined algorithmically, just like predicate
transformer semantics (e.g., strongest post-condition).  Here, a state
transformer interprets a SQL statement in the set domain, taking
advantage of the fact that the database is essentially a set of
tuples, and a SQL statement is a transformer over these sets.  The
benefit of this approach is that low-level loops can now be
substituted with higher-order combinators that automatically lift the
state transformer of its higher-order argument (i.e., the loop body)
to the state transformer of the combined expression (i.e., the loop).
Thus the semantics of a \C{foreach} loop, for instance, can be
captured as a state transformer, where the state is a set, and the
transformation defines a bind operation. We illustrate this intuition
on a simple example.

>>>>

We present our ideas in the context of an embedded DSL in OCaml, which
offers a \C{DB} monad to define and compose database computations
given in terms of SQL operations.  Besides the usual \C{bind} and
\C{return} operators, the monad includes an \C{atomically} combinator
that executes a database computation as an atomic transaction and
returns the result.

@@

Fig.~\ref{fig:new_order_code} shows a simplified version of the TPC-C
\C{new\_order} transaction written in this language. TPC-C is a
widely-used and well-studied Online Transaction Processing (OLTP)
benchmark that models an order-processing system for a wholesale parts
supply business. The business logic is captured in 5 database
transactions that operate on 9 tables; \C{new\_order} is one such
transaction that uses \C{District}, \C{Order}, \C{New\_order},
\C{Stock}, and \C{Order\_line} tables. The transaction acts on the
behalf of a customer, whose id is \C{c\_id}, to place a new order for
a given set of items (\C{item\_reqs}), to be served by a warehouse
under the district identified by \C{d\_id}.  Fig.~\ref{fig:schema}
illustrates the relationship among these different tables.

The transaction manages order placement by invoking appropriate SQL
functionality, captured by various calls to functions defined by the
\C{SQL} module. All \C{SQL} operators supported by the module take a
table name (a nullary constructor) as their first argument. The
higher-order \C{SQL.select1} function accepts a boolean function that
describes the selection criteria, and returns any record that meets
the criteria (it models the SQL query \C{SELECT \ldots\xspace LIMIT
  1}). \C{SQL.update} also accepts a Boolean function (its 3$^{rd}$
argument) to select the records to be updated. Its 2$^{nd}$ argument
is a function that maps each selected record to a new (updated)
record. \C{SQL.insert} inserts a given record into the specified table
in the database.

@@

\begin{figure}[t]
  \centering
	\begin{subfigure}{0.4\textwidth}
		\includegraphics[width=0.9\textwidth]{Figures/schema1}
    \caption{A valid TPC-C database. The only existing order belongs
      to the district with \C{d\_id}=11. Its id (\C{o\_id}) is one
      less than the district's \C{d\_next\_o\_id}, and its order
      count (\C{o\_ol\_cnt}) is equal to the number of order line
      records whose \C{ol\_o\_id} is equal to the order's id.  }
		\label{fig:tpcc_db1}
	\end{subfigure}
	\begin{subfigure}{0.52\textwidth}
		\includegraphics[width=0.9\textwidth]{Figures/schema2}
    \caption{The database in Fig.~\ref{fig:tpcc_db1} after correctly
      executing a \C{new\_order} transaction. A new order record is
      added whose \C{o\_id} is equal to the \C{d\_next\_o\_id} from
      Fig.~\ref{fig:tpcc_db1}. The district's \C{d\_next\_o\_id} is
      incremented. The order's \C{o\_ol\_cnt} is 2, reflecting the
      actual number of order line records whose \C{ol\_o\_id} is equal
      to the order's id (2).}
		\label{fig:tpcc_db2}
	\end{subfigure}

@@

Because the aforementioned reasoning is reasonably simple to perform
manually, verifying the soundess of TPC-C's consistency conditions
would appear to be feasible.  Serializability aids the tractability of
verification by preventing any interference among concurrently
executing transactions while the \C{new\_order} transaction executes.
Under weak isolation\footnote{Weak isolation does not violate
  atomicity as long as the witnessed effects are those of committed
  transactions}, however, interferences of various kinds are
permitted, leading to executions superficially similar to executions
permitted by concurrent (racy) programs~\cite{GHE15,HPQ+15}.
\begin{wrapfigure}{l}{.4\textwidth}
\includegraphics[scale=0.45]{Figures/motiv-eg-1-b}
\caption{\small An RC execution involving two instances ($T_1$ and
  $T_2$) of the \C{new\_order} transaction depicted in
  Fig.~\ref{fig:new_order_code}. 
  Both instances read the \C{d\_id} \C{District} record concurrently,
  because neither transaction is committed when the reads are
  executed.  The subsequent operations are effectively sequentialized,
  since $T_2$ commits. Nonetheless, reading the same value for
  \C{d\_next\_o\_id} prompts both instances to add \C{Order} records
  with same ids, which in turn triggers a violation of TPC-C's
  consistency condition.}
\label{fig:new_order_exec}
\end{wrapfigure}
To illustrate, consider the behavior of the \C{new\_order} transaction
when executed with a \emph{Read Committed} (RC) isolation level, the
default isolation level in 8 of the 18 databases studied
in~\ref{bailishat}.  An RC transaction is isolated from \emph{dirty
  writes}, i.e., writes of uncommitted transactions, but is allowed to
witness the writes of concurrent transactions as soon as they are
committed. Thus, with two concurrent instances of the \C{new\_order}
transaction (call them $T_1$ and $T_2$), both concurrently placing new
orders for different customers under the same district (\C{d\_id}), RC
isolation allows the execution shown in
Fig.~\ref{fig:new_order_execs}.

@@

Notably, this example does not exhibit any of the anomalies that
characterize RC isolation~\cite{berenson}. For instance, there are no
\emph{lost writes}; both concurrent transactions' writes are present
in the final state of the database. Indeed, program analyses that aim
to determine appropriate isolation by checking for possible
manifestations of RC-induced anomalies would fail to identify grounds
for promoting the isolation level of \C{new\_order} to something
stronger.  Yet, if we take the semantics of the application into
account, it is quite clear that RC is not an appropriate isolation
level for \C{new\_order}.

While reasoning in terms of anomalies is cumbersome and inadequate,
reasoning about weak isolation in terms of low-level
traces~\cite{adyaphd,gotsmanconcur15} on memory read and write actions
can complicate high-level reasoning.  A possible alternative would
interleave weak isolation implementation details within the program,
yielding a (more-or-less) conventional concurrent program that can be
then subject to classical concurrent verification methods.
Considering the size and complexity of real-world transaction systems,
this strategy is unlikely to scale.

In this paper, we adopt a different approach that \emph{lifts}
isolation semantics (\emph{not} their implementations) to the
application layer, providing a principled framework to simultaneously
reason about application invariants and isolation properties.  To
illustrate this idea informally, consider how we might verify that
\C{new\_order} is sound when executed under \emph{Snapshot Isolation}
(SI), a stronger isolation level than RC. Snapshot isolation allows
transactions to be executed against a private snapshot of the
database, thus admitting concurrency, but it also requires that there
not be any write-write conflicts among concurrent transactions when
they commit. Write-write conflicts can be eliminated in various ways,
e.g., through conflict detection followed by a rollback, or through
exclusive locks, or a combination of both.  For instance, one possible
implementation of SI, close to the one used by
PostgreSQL~\cite{postgres-ssi}, executes a transaction against its
private snapshot of the database, but obtains exclusive locks on the
actual records in the database before performing writes. A write is
performed only if the record that is to be written has not already
been updated by a concurrent transaction. Conflicts are resolved by
abort and roll back.

%% This implementation is demonstrated for a simple
%% database with three records, $a$, $b$, and $c$ in
%% Fig.~\ref{fig:RR-postgres}.

%% \begin{figure}[t]
%% \includegraphics[scale=0.5]{Figures/RR-postgres}
%% \caption{Database state transitions corresponding to an execution of
%%   an SI transaction $T_1$ on a PostgreSQL-like store. $T_2$ and $T_3$
%%   are concurrent transactions. }
%% \label{fig:rr-postgres}
%% \end{figure}
%% \begin{figure}[t]
%% \includegraphics[scale=0.5]{Figures/RR-abstract}
%%   \caption{An abstract execution that includes the concrete execution
%%   shown in Fig.~\ref{fig:rr-postgres}. It has no locks or snapshots,
%%   admits fewer interleavings, yet results in the same post-state.  }
%% \label{fig:rr-abstract}
%% \end{figure}

Implementations of SI on real databases (e.g., PostgreSQL) are
complicated, often running into thousands of lines of code.
Nonetheless, the semantics of SI, in terms of how it effects
transitions on the database state, can be captured in a fairly simple
model. First, effects induced by one transaction are not visible to
another concurrently executing one (call it $T$), during $T$'s
execution. Thus, from $T$'s perspective, the global state does not
change during its execution.  More formally, for every operation
performed by $T$, the global state $T$ witness before ($\stg$) and
after ($\stg')$ executing the operation is the same ($\stg'=\stg$).
After $T$ finishes execution, it commits its changes to the actual
database, which may have already incorporated the effects of concurrent
transactions. In executions where $T$ successfully commits, concurrent
transactions are guaranteed to not be in write-write conflict with
$T$. Thus, if $\stg$ is the global state that $T$ witnessed when it
finished execution (the snapshot state), and
$\stg'$ is the state to which $T$ commits, then the difference between
$\stg$ and $\stg'$ should not result in a write-write conflict with $T$. To
concretize this notion, let the database state be a map from
transaction variables to values, and let $\stl$ denote a
transaction-local log that maps the variables being written to their
updated values.  The absence of write-write conflicts between $T$
and the diff between $\stg$ and $\stg'$ can be expressed as: $\forall
x\in\mathit{dom}(\delta)$, $\stg'(x) = \stg(x)$.  In other words, the
semantics of SI can be captured as an axiomatization over transitions
of the database state ($\Delta \longrightarrow \Delta'$) during the
lifetime an SI transaction ($T$):
\begin{itemize}
  \item While $T$ executes, $\Delta' = \Delta$.
  \item After $T$ finishes execution, but before it commits its local
    state $\delta$, $\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.
\end{itemize}

@@

\noindent This simple characterization of SI isolation allows us to
verify the consistency conditions associated with the \C{new\_order}
transaction.  First, since the database doesn't change ($\Delta' =
\Delta$) during the execution of the transaction's body, we can reason
about \C{new\_order} as though it executed in complete isolation until
its commit point, leading to a verification process similar to what
would been applied when reasoning about serializability.  When
\C{new\_order} finishes execution, but before it commits, the SI
axiomatization shown above requires us to consider global state
transitions $\stg \longrightarrow \stg'$ that do not include changes
to the records ($\stl$) written by \C{new\_order}, i.e.,
$\forall(x\in\delta).~\Delta'(x) = \Delta(x)$.  The axiomatization
precludes any execution in which there may concurrent updates to
shared table fields (e.g., \C{d\_next\_o\_id} on the same \C{District}
table), but does not prohibit interferences that write to different
tables, or different records in the same table.  We need to reason
about the safety of such interferences with respect to the
transaction's consistency invariants.

%% Since concurrent
%% \C{new\_order} transactions that write to the same \C{District} record
%% modify the record by incrementing its \C{d\_next\_o\_id} field, their
%% transitions do not satisfy the necessary condition imposed by the SI
%% axiomatization, allowing us to rightfully ignore their interference.
%% This leaves the interference due to concurrent \C{new\_order}
%% transactions that write to a different district record.  Fortunately,
%% such interference does not prompt \C{new\_order} to violate the
%% invariant, which can be established by adopting a reasoning similar to
%% the one used on concurrent program logics, such as the Rely-Guarantee
%% logic, customized for the database programs.

@@

%% The second observation that informs our approach is one that pertains
%% to automation. Program verification, even when machine-aided, often
%% entails significant annotation burden in the form of intermediary
%% assertions and loop invariants required to prove a program correct.
%% This is certainly true for concurrent program logics, such as
%% Rely-Gurantee, which extend Hoare logic with additional artifacts and
%% where (stable) intermediary assertions and loop invariants remain a
%% major source of annotation burden.

To do so, we observe that a relational database is a significantly
simpler abstraction than shared memory. There are no pointers, linked
data structures, or aliasing.  Although a database essentially
abstracts a mutable state, the state is affected through a
well-defined fixed number of interfaces (SQL statements), each tagged
with a logical formula describing what records are accessed and
updated.

This observation leads us away from thinking of database transactions
as concurrent imperative programs.  Instead, we see value in viewing
them as essentially functional computations that manage database state
monadically.  By doing so, we can formulate the semantics of database
operations as state transformers that relate the operation's pre- and
post-states.  A state transformer semantics can be defined
algorithmically, just like classical predicate transformer semantics
(e.g., strongest post-condition).  Here, a transformer interprets a
SQL statement in the set domain, modeling the database as a set of
tuples, and a SQL statement a function over this set.  Among other
things, one benefit of this approach is that low-level loops can now
be substituted with higher-order combinators that automatically lift
the state transformer of its higher-order argument (i.e., the loop
body) to the state transformer of the combined expression (i.e., the
loop).  We illustrate this intuition on a simple example.

%% Thus the semantics of a \C{foreach} loop, for instance, can be
%% captured as a state transformer, where the state is a set, and the
%% transformation defines a bind operation.
